---
title: "EDA for a Soccer Data"
author: "Francis Addae"
date: "Last Edited: `r format(Sys.time(), '%B %d, %Y')` "
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    df_print: paged
    toc: yes
fontsize: 11pt
---

```{r setup, include=FALSE, warning=FALSE,error=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(forcats)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6, warning = FALSE)
```

```{r}
#setting theme# 
mytheme <- theme_bw() +  theme(
  plot.title = element_text(color = "black", size = 15, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(color = "black", face = "italic", hjust = 0.5),
  plot.caption =  element_text(color = "red", face = "italic")
)

pietheme <- theme_void() +  theme(
  plot.title = element_text(color = "black", size = 15, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(color = "black", face = "italic", hjust = 0.5),
  plot.caption =  element_text(color = "red", face = "italic")
)
```

#Summary
            This is an Exploratory Data Analysis report which focuses on exploration and gaining insights.

#Libraries 
Below are the libraries used in this report. 
* **dpylr** : For data query, wrangling, manipulation and transformation.
* **ggplot2**: For data visualization and info messaging 
* **forcats**: For categorical variable manipulation and wrangling
* **lubridates**: For date-time variable analysis and reporting

#Data 
This dataset is a soccer data obtained from [kaggle](https://www.kaggle.com/isabey/soccer-results-csv?select=results.csv).It is the collection of all intenational games played from the inception of association football or soccer. Association football's history and understanding can be found on [Wikipedia](https://en.wikipedia.org/wiki/History_of_association_football) 
```{r}
soc_1800_2020 <- read.csv('soccer_Data.csv')
shape <- dim(soc_1800_2020)
numCol <- ncol(soc_1800_2020)
```

##Data Summary and Cleaning

###Summary
This data is the collection of games played between nations from 1800's to 2020. It data has a dimension of `r shape`; such records are attributed to `r numCol`. Each record is match between two teams. Any recorded game has a unique home and away team. The attributes are: 
* **date**:  Date when a game was played. This isnt a date-time variable.
* **home_team**: Name of the home team 
* **away_team**: Name of the away team
* **home_score**: Score made by the home team in that particular match
* **away_score**: Score made by the away team in that particular match
* **tournament**: The type of tournamnet in whcih the game was held. 
* **city**: The city where the game was officiated.
* **country**: The country in which the city is located.
* **neutral**: The advantageous venue of each match.Think of it as home_court advantage
* **result**:  The final result of the game. 

Although these attributes are great, the dataset could be served with attributes like; temperature, time, attendance, stadium name etc, which can give great insght as to how the overall game look like data-wise.

###Cleaning 
```{r}
nulls <- sum(is.na(soc_1800_2020))
```
The dataset has `r nulls` null(s) values. Despite this, the dataset could use further cleaning if need be. For example, instead of have both the *city* and *country* as a seperate entity, they can be combined as a unique value. This can shorten the dimension of the data. A lot of cleaning can come in place after the formatting of the dataset. 



#Data Formatting 

```{r}
#Date --> year-month-day formart 
#Away & Home team are in factors(categorical variable)
#Away & Home scores in integers 
#City & Country is the location where a game was played
#Neutral --> if a game was played on a home or away team.
#Result --> Which team won during the match? homeWin, awayWin or draw 

soc_1800_2020$date <-  ymd(soc_1800_2020$date)
soc_1800_2020$home_team <- as.factor(soc_1800_2020$home_team)
soc_1800_2020$away_team <- as.factor(soc_1800_2020$away_team)
soc_1800_2020$home_score <- as.integer(soc_1800_2020$home_score)
soc_1800_2020$away_score <- as.integer(soc_1800_2020$away_score)
soc_1800_2020$tournament <- as.factor(soc_1800_2020$tournament)
soc_1800_2020$city       <- as.factor(soc_1800_2020$city)
soc_1800_2020$country    <- as.factor(soc_1800_2020$country)
soc_1800_2020$neutral    <- as.factor(soc_1800_2020$neutral)
soc_1800_2020$result     <- as.factor(soc_1800_2020$result) 
```


#Objective

```{r}
####  Exploratory Data Analysis #### 
#How many games were played each year?
ans <- soc_1800_2020 %>% group_by(year(date)) %>% 
  summarize(ct = n())
ggplot(ans, aes(`year(date)`, ct)) + geom_point() + geom_smooth(method = 'lm') + 
  labs(x = 'Year', y= 'gamesPlayed', title = 'Number of games played each year')
```

```{r}
#What are the top 10 home teams in soc_1800_2020cer?
ans <- soc_1800_2020 %>% select(home_team) %>% group_by(home_team) %>% summarize(ct = n()) %>% 
  arrange(desc(ct)) %>% top_n(10) 
ggplot(ans, aes(home_team, ct, size = ct)) + geom_point() + 
  labs(x = 'home team', y = '# of games',title = 'Top 10 hometeams', size = 'Games Played')
```

```{r}
#What are the top 10 away teams in soc_1800_2020cer? 
ans <- soc_1800_2020 %>% group_by(away_team) %>% summarize(ct = n()) %>% 
  arrange(desc(ct)) %>% top_n(10) 
ggplot(ans, aes(away_team, ct, size = ct)) + geom_point() + 
  labs(x = 'away team', y = '# of games',title = 'Top 10 awayteams')
```

```{r}
#Distribution of home scores?
breaks <- c(0, min(soc_1800_2020$home_score) + 1, mean(soc_1800_2020$home_score) + 5,max(soc_1800_2020$home_score))
ggplot(soc_1800_2020, aes(home_score)) + geom_histogram(breaks = breaks) + 
  labs(x = 'home_scores', y = 'scores count', title = 'Distribition of home scores')
```

```{r}
#Distribution of away scores?
breaks <- c(0, min(soc_1800_2020$away_score) + 1, mean(soc_1800_2020$away_score) + 5,max(soc_1800_2020$away_score))
ggplot(soc_1800_2020, aes(away_score)) + geom_histogram(breaks = breaks) + 
  labs(x = 'away_scores', y = 'scores count', title = 'Distribition of away scores')
```

```{r}
#What's the total goals scored each year through out the dataset? 
ans <- soc_1800_2020 %>% group_by(year(date)) %>% summarise(hmGoals = sum(home_score), ayGoals = sum(away_score)) %>% 
  select(yr = `year(date)`, hmGoals, ayGoals)
ggplot(ans, aes(x = yr)) + geom_line(aes(y=hmGoals), col = 'darkred') + 
  geom_text(aes(x = 1990, y = 1550, label = 'home goal')) + geom_line(aes(y=ayGoals),col = 'darkblue') + 
  geom_text(aes(x = 2005, y = 700, label = 'away goal')) + 
  labs(x = 'year', y = 'Total Goals Scored',title = 'Summaztion of Goals throughout each year')
```

```{r}
#What are the top 10 tournaments in played? 
ans <- soc_1800_2020 %>% group_by(tournament) %>% summarise(ct = n()) %>% 
  arrange(desc(ct)) %>% top_n(25)
ggplot(ans, aes(reorder(tournament, -ct), ct)) + geom_col() + coord_flip() + 
  labs(x = 'tournament', y = '# of games', title = 'Top 25 tournaments')
```

```{r}
#create a geolocation and see what are the top 25 locations where a game was hosted? 
ans <- soc_1800_2020 %>% mutate(location = paste0(city,',',country)) %>% group_by(location) %>% 
  summarize(numUse = n()) %>% arrange(desc(numUse)) %>% top_n(25) 
ggplot(ans,aes(location,numUse)) + geom_segment(
  aes(x =location, xend = location, y = 100, yend=numUse), color = 'skyblue') + 
  geom_point(color = 'blue', size = 4, alpha = 0.6) + coord_flip() +
  labs(x = 'venue', y = '# of games', title = 'Top 25 locations where games were hosted')
```

```{r}
#Distribution of Neutral zones through each year 
ans <- soc_1800_2020 %>% group_by(neutral, year(date)) %>% summarize(ct = n()) %>% 
  select(neutral, yr = `year(date)`, ct)
ggplot(ans, aes(yr, ct, color = neutral)) + geom_line() + 
  labs(x = 'Date(year)', y = '# of neutral', title = 'Games played at home/away feilds vs neutral fields')
```

```{r}
#Results throughout each year? 
ans <- soc_1800_2020 %>% group_by(result, year(date)) %>% summarize(ct = n()) %>% 
  select(result, yr = `year(date)`, ct)
ggplot(ans, aes(yr, ct, color = result)) + geom_line() + 
  labs(x = 'Date(year)', y = '# of neutral', title = 'Game results during each year') 
```

```{r}
#Objective: Do home_teams that perform better in the late 1990's perform the same as the early 2000's?  
#Looking records from Jan,1990 to December, 1999 
soc_1990_1999 <-soc_1800_2020 %>% filter(date >= '1990-01-01' & date <= '1999-12-31') %>%
  select(date, home_team, home_score, tournament, city, country, result) 
```

```{r}
#What's the overall result for each each year?
ans <- soc_1990_1999 %>% group_by(year(date),result) %>% tally(sort = TRUE) %>% 
  select(yr = `year(date)`, result, pnt = n)
ggplot(ans, aes(yr, pnt, col = result)) + geom_line() + 
  labs(x = 'Year', y = 'Results Count', title = 'Result from matches from 1990-1999')
```


```{r}
#What's the summary description of both home goals throughout each year?
ggplot(soc_1990_1999, aes(factor(year(date)), home_score)) + geom_boxplot() + 
  labs(x = 'Year', y = 'Scores', title = 'Descriptive Summary description of Home gaols from 1990-1999')
```

```{r}
#What's the best location for a team to play in each year? 
ans <- soc_1990_1999 %>% mutate(location = paste(city,',',country)) %>% group_by(year(date), location) %>% 
  filter(result == 'homeWin') %>% tally(sort = TRUE) %>% select(location, yr = `year(date)`, pnt = n) %>% 
  group_by(yr) %>% top_n(10, location)
ggplot(ans, aes(reorder(location,pnt), pnt, fill = factor(yr))) + geom_col() + coord_flip() +
  labs(x = 'Location: city,country', y = '# of homewins', fill = 'Year',
       title = 'Top 10 locations for home teams to win each year')
```


```{r}
#What's the win percentage each year? 
#We need the number of games played each year  and divide that by the number of homewins each year
#Count of records each year 
x <- soc_1990_1999 %>% group_by(year(date)) %>% tally() %>% select(yr = `year(date)`, pnt = n)

y <- soc_1990_1999 %>% group_by(result, year(date)) %>% tally() %>% 
  filter(result == 'homeWin') %>% select(yr = `year(date)`, homeWin = n,result)

```

```{r}
ans<- x %>% inner_join(y, by = 'yr') %>% 
  mutate(winPec = round(homeWin/pnt * 100), gamesRatio = round(pnt/sum(pnt) * 100)) %>% 
  select(yr, homeWin, winPec, gamesRatio, pnt)
ans
ggplot(ans, aes(yr)) + geom_bar(aes(y = homeWin), stat = 'identity') + 
  geom_text(aes(y = homeWin+20, label = paste0(winPec,'%'))) + geom_line(aes(y = pnt), color = 'darkred') + 
  geom_text(aes(y = pnt+10, label = paste0(gamesRatio,'%'))) + 
  labs(title = 'Wins Percentage through 1990-1999', x = 'Year', y = 'gamesPlayed')
```


```{r}
#Looking at records from Januaray,2000-December,2010
soc_2000_2020 <- soc_1800_2020 %>% filter(date >= '2000-01-01' & date <= '2010-12-31') %>%
  select(date, home_team, home_score, tournament, city, country, result) 
soc_2000_2020
```

```{r}
#What's the overall result for each each year?
ans <- soc_2000_2020 %>% group_by(year(date),result) %>% tally(sort = TRUE) %>% 
  select(yr = `year(date)`, result, pnt = n)
ggplot(ans, aes(yr, pnt, col = result))+ geom_line() + 
  labs(x = 'Year', y = 'Result', title = 'Game results from 2000-2010') 
```

```{r}
#What's the summary description 
ggplot(soc_2000_2020, aes(factor(year(date)), home_score)) + geom_boxplot() + 
  labs(x = 'Year', y = 'Scores',title = 'Descriptive Summary description of Home gaols from 2000-2010')
```

```{r}
#What's the best location for a team to play in each year? 
ans <- soc_2000_2020 %>% mutate(location = paste(city,',',country)) %>% group_by(year(date), location) %>% 
  filter(result == 'homeWin') %>% tally(sort = TRUE) %>% select(location, yr = `year(date)`, pnt = n) %>% 
  group_by(yr) %>% top_n(10, location)
ggplot(ans, aes(reorder(location,pnt), pnt, fill = factor(yr))) + geom_col() + coord_flip() +
  labs(x = 'Location: city,country', y = '# of homewins', fill = 'Year',
       title = 'Top 10 locations for home teams to win each year')
```

```{r}
#What's the win percentage each year? 
x <- soc_2000_2020 %>% group_by(year(date)) %>% tally() %>% select(yr = `year(date)`, pnt = n)

y <- soc_2000_2020 %>% group_by(result, year(date)) %>% tally() %>% 
  filter(result == 'homeWin') %>% select(yr = `year(date)`, homeWin = n)

ans<- x %>% inner_join(y, by = 'yr') %>% 
  mutate(winPec = round(homeWin/pnt * 100), gamesRatio = round(pnt/sum(pnt) * 100)) %>% 
  select(yr, homeWin, winPec, gamesRatio, pnt)
ans
ggplot(ans, aes(yr)) + geom_bar(aes(y = homeWin), stat = 'identity') + 
  geom_text(aes(y = homeWin+20, label = paste0(winPec,'%'))) + geom_line(aes(y = pnt), color = 'darkred') + 
  geom_text(aes(y = pnt+15, label = paste0(gamesRatio,'%'))) + 
  labs(title = 'Wins Percentage through 2000-2010', x = 'Year', y = 'gamesPlayed')
```



Interesting Insight/Questions: 
1.As the games home teams play, thier overall win percentage goes up. 
Therefore, there is a small correlation
2.Total home team wins rose during the middle of the 1990's but differs quite alot during the 2000's
3.Is there a correlation between games played and goals scored? Can a difference in goals shed a light  
4.What there specific years where homeWins excited awayWins. Which teams won the most?  

```{r}
#Correlation between goals and result.{Hypothesis: More home goals leads to more home win?} 
#focusing on on the previous two decades { 1990-2010} 
dt <- soc_1990_1999 %>% full_join(soc_2000_2020) %>% mutate(location = paste(city,',',country)) %>% 
  select(date, team = home_team, goals = home_score, tournament, location, result)
dt$location <- as.factor(dt$location)
summary(dt)
```

```{r}
#General case
ggplot(dt %>% group_by(year(date)) %>% summarize(hmGoal = sum(goals), games = n()) %>% 
  select (yr = `year(date)`, hmGoal, games), aes(yr)) + geom_line(aes(y=hmGoal), col = 'darkred') + 
  geom_col(aes(y=games), fill= 'darkblue')
```

```{r}
#When home team won 
ggplot(dt %>% group_by(year(date)) %>% filter(result == 'homeWin') %>% 
  summarize(hmGoal = sum(goals), games = n()) %>% select (yr = `year(date)`, hmGoal, games)
       ,aes(yr)) + geom_line(aes(y=hmGoal), col = 'darkred') + geom_col(aes(y=games), fill= 'darkblue')
```

```{r}
#Those doesnt really answer the question. Let's see the average( total goals/ total games) per year when the
ggplot(dt %>% group_by(year(date)) %>%  filter(result == 'homeWin') %>% 
  summarize(avgGoal = mean(goals)) %>% select(yr = `year(date)`, avgGoal), 
  aes(yr, avgGoal)) + geom_col() 

#It seems more goals doesn't grarantee more future success in wins for any home team.
```

```{r}
#Check if continents can factor in wins for hometeams
continents <- read.csv('countries-continents.csv')
continents
```

```{r}
#General case
ggplot(dt %>% left_join(continents, by = c('team' = 'Country')) %>% filter(Continent != 'NA') %>% 
  group_by(Continent) %>%  summarize(pnt = sum(goals), gamesPlayed = n()), aes(Continent)) + 
  geom_col(aes(y = pnt), fill = 'red') + geom_col(aes(y=gamesPlayed), fill = 'blue') + 
  labs(caption = "Total Goals Scored in Red, GamesPlayed in Blue")
```

```{r}
#Home team wining 
ggplot(dt %>% left_join(continents, by = c('team' = 'Country')) %>% 
         filter(Continent != 'NA' & result == 'homeWin') %>% group_by(Continent) %>%  
         summarize(pnt = sum(goals), gamesPlayed = n()), aes(Continent)) + 
  geom_col(aes(y = pnt), fill = 'red') + geom_col(aes(y=gamesPlayed), fill = 'blue') + 
  labs(caption = "Total Goals Scored in Red, GamesPlayed in Blue")
```


It seems more Asia teams are more likely to score as much goals as European team at home. Conclusively,we can say more home goals some home teams benifits a,ot from scores more home goals; especially Asian and European teams 

#Classification of Games. 
The most important thing is wether a teams wins coniceds with the field advantage.Field advantage is when a home or away team gets to play on thier own turf. Can a win be predicted based on filed neutral. Neutrality is when True --> When a games is played on a Neutral field; FALSE when not 

Let's add the continents to the data. This will be done by using the country at which the game was played.Reason why:Is the simplest way to explain aggragates and we dont need to factor in which teams gets the label. 


```{r}
#Merging the continents in the dataset
countries_and_continents <- continents %>% mutate(
  country = ifelse(nchar(Country) <= 2, case_when(Country == 'CZ'~ 'Czech Republic',
                                                   Country == 'US' ~ 'United States',
                                                   TRUE ~ as.character(Country)), Country)) %>% 
  rename(continent = Continent) %>%  select(country, continent)


dt <- soc_1800_2020 %>% full_join(countries_and_continents, by = 'country')

```

```{r}
#Lets do a little exploration with the data?

#What's the total games played on each?
ggplot(dt %>% group_by(continent) %>% tally(), aes(continent, n)) + geom_col() + 
  labs(x = 'Continents', y = '# of games', title = 'Number of games played on each continent')
```

```{r}
#What's the total games played on each continent by neutrality?
ggplot(dt %>% group_by(continent,neutral) %>% tally(), aes(continent, n, fill = neutral)) + 
  geom_col(position = 'dodge') + labs(x = 'Continents', y = '# of games', 
                            title = 'Number of games played on each continent')
```

```{r}
#What's the rolling average of games played each year by continent? 
ans <- dt %>% group_by(year(date), month(date),continent) %>% summarize(gamesPlayed = n()) %>% 
  select(year = `year(date)`, month = `month(date)`, gamesPlayed, continent) %>% arrange(year, month) %>% 
  mutate(rollingAvgGames = cumsum(gamesPlayed)/ row_number(),
         date = make_date(year = year, month = month)) %>% 
  select(date,gamesPlayed, rollingAvgGames, continent)
ans 
ggplot(ans, aes(date, gamesPlayed)) + geom_line() + facet_grid(continent~.) + 
  scale_x_date(date_minor_breaks = "10 years") + 
  labs(y ='# of games', title = 'Total games played in each continent') 
```

```{r}
ggplot(ans, aes(date, rollingAvgGames)) + geom_line() + facet_grid(continent~.) + 
  scale_x_date(date_minor_breaks = "10 years") + 
  labs(y ='# of games', title = 'Moving averages of games played on each continent') 
```



```{r}
#Rolling Averages of wins per year for each continent
ans <- dt %>% group_by(continent, year(date), result) %>% filter(result != 'draw') %>%  
  summarize(totalWins = n()) %>%  mutate(avgWins = cumsum(totalWins)/ row_number()) %>% 
  select(year = `year(date)`,continent,totalWins, avgWins,result) %>% ungroup()
ans
ggplot(ans, aes(year, totalWins, col = result, size = avgWins)) + geom_jitter() + facet_wrap(~continent) +
  labs(x = 'year', y = 'total Wins per year', col = 'Type of Wins', size = 'rolling Win Averages',
       title = "Can the continent in which a game was played affects it's win totals?") 
```



```{r}
ggplot(ans, aes(year, totalWins, col = result, size = avgWins)) + geom_jitter() + facet_wrap(~continent) +
  labs(x = 'year', y = 'total Wins per year', col = 'Type of Wins', size = 'rolling Win Averages',
       title = "Can the continent in which a game was played affects it's win totals?",
       subtitle = "Smoothing curve: Linear Model") + geom_smooth(method = 'lm')
```


#Observation: 
We previously know that Away teams won most games in each tournament.By using the rolling average of wins per each continent. we can see the trends within each facets. 
1. Games played on European, African and North America soils tend to have the most away wins. Meaning: Home teams tend to lost when thier play awayteam's on thier own turf.
2. Oceaniaic and South America hosted games tend to have most home game wins but ther win progression isnt as strongly correlated as European teams. 
3. Asia hosted games have a bottleneck in terms in both home and away wins. 
4. Oceaniaic and South America games have lower wining averages, there by reducing the win averages. 
 


```{r}
#Are they any changes within the rolling wins each year?
 ggplot(ans %>% group_by(result) %>% mutate(winsChanges = avgWins - lag(avgWins)) %>% 
  arrange(desc(year)), aes( year, winsChanges)) + geom_line() + facet_wrap(~continent)
```
